---
title: " "
date: " "
bibliography: citations.bib
nocite: "@*"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 1
    theme: flatly
    embed-resources: true
    page-layout: full
    code-fold: true
    number-sections: false
    html-math-method: katex
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
    geometry:
      - top=25mm
      - left=20mm
      - right=20mm
      - bottom=25mm
    include-in-header:
      text: |
        \usepackage{fancyhdr}
        \usepackage{titling}
        \usepackage{tabu}
        \usepackage{booktabs}
        \usepackage{sectsty}
        \usepackage[skins,breakable]{tcolorbox}
        \usepackage[scaled]{helvet}
        \usepackage{url}
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \renewcommand\familydefault{\sfdefault}
        \sectionfont{\color[rgb]{0,0.3,0.6}\huge\bfseries}
        \subsectionfont{\color[rgb]{0,0.3,0.6}\large\bfseries}
        \pagestyle{fancy}
        \fancyhf{}
        \renewcommand{\headrulewidth}{0.4pt}
        \renewcommand{\footrulewidth}{0.4pt}
        \fancyhead[R]{\includegraphics[height=0.5cm]{milou_logo.png}}
        \fancyhead[L]{\small\textit{Specimen Status}}
        \fancyfoot[L]{\small milou Analysis Report | Pipeline v1.2.0 | UUID: UUID_PLACEHOLDER}
        \fancyfoot[R]{\thepage}
        \fancypagestyle{plain}{
          \fancyhf{}
          \renewcommand{\headrulewidth}{0pt}
          \renewcommand{\footrulewidth}{0.4pt}
          \fancyfoot[L]{\small milou Analysis Report | Pipeline v1.2.0 | UUID: UUID_PLACEHOLDER}
          \fancyfoot[R]{\thepage}
        }
        % Interpretation Note Box
        \newtcolorbox{reportnote}{colback=blue!5!white,colframe=blue!75!black,title=Interpretation Note,fonttitle=\bfseries}
        \newtcolorbox{technicalwarning}{colback=red!5!white,colframe=red!75!black,title=Technical Data Warning,fonttitle=\bfseries}
execute:
  echo: false
  warning: false
  message: false
params:
  mode: "clinical"
  sample_metadata: "sample_metadata.tsv"
  qc_summary: "qc_summary.tsv"
  dmr_summary: "dmr_summary.tsv"
  gene_summary: "gene_prioritized.csv"
  disease_summary: "disease_enrichment.csv"
  pathway_summary: "go_results.csv"
  kegg_summary: "kegg_results.csv"
  run_info: "run_info.json"
---

```{r setup}
# Using base R instead of tidyverse to ensure compatibility with all container environments
library(glue)
library(knitr)
library(kableExtra)
library(jsonlite)

# Gracefully try to load or dynamically install DT if missing
if (!requireNamespace("DT", quietly = TRUE)) {
  tryCatch({
    local_lib <- file.path(getwd(), "R_libs")
    if (!dir.exists(local_lib)) dir.create(local_lib, recursive = TRUE)
    .libPaths(c(local_lib, .libPaths()))
    options(repos = c(CRAN = "https://packagemanager.posit.co/cran/__linux__/noble/2026-05-01"))
    install.packages("DT", lib = local_lib, dependencies = FALSE, quiet = TRUE)
  }, error = function(e) {
    message("Failed to install DT; falling back to static tables.")
  })
}
has_dt <- requireNamespace("DT", quietly = TRUE)
if (has_dt) {
  library(DT)
}

# Robust error handling for file reading
read_tsv_safe <- function(file_path) {
  if(!is.null(file_path) && file.exists(file_path)) {
    if (grepl(".csv$", file_path)) {
        return(read.csv(file_path))
    } else {
        return(read.delim(file_path, sep="\t"))
    }
  } else {
    return(data.frame())
  }
}

# Helper to find plots (since names have prefixes) and copy them locally for pdf/latex safety
find_plot <- function(dir_path, suffix) {
    if (!dir.exists(dir_path)) return(NULL)
    files <- list.files(dir_path, pattern = suffix, full.names = TRUE)
    if (length(files) > 0) {
        local_name <- basename(files[1])
        file.copy(files[1], local_name, overwrite = TRUE)
        return(local_name)
    }
    NULL
}

# Load data gracefully
meta <- read_tsv_safe(params$sample_metadata)
qc <- read_tsv_safe(params$qc_summary)
dmr <- read_tsv_safe(params$dmr_summary)
gene <- read_tsv_safe(params$gene_summary)
disease <- read_tsv_safe(params$disease_summary)
pathways <- read_tsv_safe(params$pathway_summary)
kegg <- read_tsv_safe(params$kegg_summary)

# Get results_dir from one of the params
res_dir <- dirname(params$dmr_summary)

if (file.exists(params$run_info)) {
    run_info <- fromJSON(params$run_info)
} else {
    run_info <- list()
}

# Pull software versions into a separate variable for easier access in appendix
software_versions <- if (!is.null(run_info$software_versions)) run_info$software_versions else list()

# Helper for HTML/LaTeX-safe tables
kable_report <- function(df, caption = "", font_size = 10) {
  if (nrow(df) == 0) return(cat("No data available.\n"))
  
  if (knitr::is_html_output()) {
    # Interactive HTML Table
    colnames(df) <- gsub("\\.", " ", colnames(df))
    colnames(df) <- gsub("_", " ", colnames(df))
    if (exists("has_dt") && has_dt) {
      return(DT::datatable(df, 
                           caption = caption, 
                           extensions = 'Buttons',
                           options = list(
                             dom = 'Bfrtip',
                             buttons = c('csv', 'excel', 'copy'),
                             pageLength = 10,
                             scrollX = TRUE
                           ),
                           rownames = FALSE))
    } else {
      # Graceful fallback if DT is not available
      kb <- kable(df, caption = caption)
      return(kable_styling(kb, full_width = TRUE, 
                           bootstrap_options = c("striped", "hover", "condensed")))
    }
  } else {
    # Static PDF Table
    colnames(df) <- gsub("\\.", " ", colnames(df))
    colnames(df) <- gsub("_", " ", colnames(df))
    colnames(df) <- gsub("%", "\\\\%", colnames(df))
    
    char_cols <- sapply(df, is.character)
    if (any(char_cols)) {
      df[char_cols] <- lapply(df[char_cols], function(x) {
        x <- gsub("_", "\\\\_", x)
        x <- gsub("%", "\\\\%", x)
        x
      })
    }
    
    kb <- kable(df, escape = FALSE, caption = caption, booktabs = TRUE)
    kable_styling(kb,
                  full_width = FALSE, 
                  bootstrap_options = c("striped", "hover", "condensed"),
                  latex_options = c("striped", "HOLD_position", "scale_down"),
                  font_size = font_size)
  }
}

n_samples <- nrow(meta)
n_dmrs <- nrow(dmr)
```
\newpage

# Executive Summary

```{r results='asis'}
if(n_samples > 0) {
    cat(glue("This report summarizes DNA methylation analysis for **{n_samples} specimens**. All analytical metrics were screened against pipeline reference standards.\n\n"))
}

if(n_dmrs > 0) {
    cat(glue("A total of **{n_dmrs} differentially methylated regions (DMRs)** were identified. Each region has been cross-referenced with functional genomic annotations and disease database associations.\n\n"))
}

if(nrow(disease) > 0 && !"Message" %in% colnames(disease)) {
    cat(glue("Top disease associations detected: **{paste(head(disease$Description, 5), collapse = ', ')}**.\n\n"))
}
```

```{r results='asis'}
narrative_field <- grep("narrative", names(run_info), value = TRUE)[1]
narrative_text <- if(!is.na(narrative_field)) run_info[[narrative_field]] else NULL
if(!is.null(narrative_text) && narrative_text != "") {
    cat("\n\n")
    cat("## Biological Interpretation\n\n")
    if (knitr::is_html_output()) {
        cat(":::{.callout-note title='Interpretation Note'}\n")
    } else {
        cat("\\begin{reportnote}\n")
    }
    cat(narrative_text, "\n")
    if (knitr::is_html_output()) {
        cat(":::\n\n")
    } else {
        cat("\\end{reportnote}\n\n")
    }
}
```

# Quality Control & Specimen Status

```{r results='asis'}
if (nrow(qc) > 0) {
  cat("## Specimen Metrics\n\n")
  cat("The following metrics summarize the sequencing and alignment quality for each specimen. Alignment rates and coverage depth are key indicators of analytical sensitivity.\n\n")
  
  kable_report(qc, caption = "Sequencing and Alignment Quality Control")
} else {
  cat("No QC data available for this run.\n")
}
```


# Differentially Methylated Regions (DMRs)

Differentially Methylated Regions (DMRs) represent genomic areas where DNA methylation levels differ significantly between experimental groups. In epigenetic research, these regions can serve as biomarkers for biological states, as epigenetic modifications often regulate gene expression without altering the underlying DNA sequence. We utilize the DSS (Dispersion Shrinkage for Sequencing) framework, which employs a beta-binomial model to account for biological variation and technical noise, providing high statistical power for detecting subtle methylation shifts.

```{r dmr-table, results='asis'}
if(n_dmrs > 0) {
    cat("## Significant Methylation Deviations\n\n")
    cat("The following table lists the top 20 most significant differentially methylated regions detected across the analytical cohort.\n\n")
    
    if (nrow(dmr) > 0) {
      dmr_top <- head(dmr[order(dmr$Min_pvalue), ], 20)
    } else {
      dmr_top <- dmr
    }
    kable_report(dmr_top, caption = "Top 20 Differentially Methylated Regions")
} else {
    cat("No significant DMRs were detected under the current significance cutoff thresholds.\n")
}
```



# Functional Gene Annotation & Prioritization

To prioritize findings with biological relevance, detected DMRs are mapped to the nearest genomic features (Promoters, Enhancers, or Gene Bodies). We apply an aggregate Prioritization Score which factors in the statistical significance (p-value), the magnitude of the methylation difference (logFC), and the proximity to transcription start sites.

The score is calculated mathematically as:
$$ \text{Rank Score} = |\log_2(\text{FC})| \times -\log_{10}(\text{FDR} + 10^{-10}) $$

## Genomic Prioritization
Ranked by aggregate prioritization score. Annotated with regional genomic context.

```{r genes-table, results='asis'}
if (nrow(gene) > 0) {
  top_genes <- gene
  if ("fdr" %in% colnames(top_genes)) {
      top_genes$fdr <- formatC(top_genes$fdr, format = "e", digits = 2)
  }
  cols_to_keep <- c("Symbol", "Region", "diff", "fdr", "Rank_Score", "Method")
  cols_present <- intersect(cols_to_keep, colnames(top_genes))
  top_genes <- head(top_genes[, cols_present, drop = FALSE], 15)
  
  kable_report(top_genes, caption = "Top 15 Genes Prioritized by Epigenetic Significance")
} else {
  cat("No gene annotations found matching the significant regions.\n")
}
```


# Pathway & Functional Interpretation

Functional enrichment analysis identifies biological processes that are over-represented among the prioritized genes. This provides a "systems-level" view of the epigenetic changes, moving beyond individual genes to identify affected cellular pathways or disease networks.

## GO Biological Process Enrichment
```{r go-results, results='asis'}
if(nrow(pathways) > 0 && !"Message" %in% colnames(pathways)) {
  cols <- intersect(colnames(pathways), c("ID", "Description", "pvalue", "p.adjust"))
  cols_present <- intersect(cols, colnames(pathways))
  top_pathways <- pathways[order(pathways$pvalue), ]
  top_pathways <- head(top_pathways[, cols_present, drop = FALSE], 10)
  print(kable_report(top_pathways, caption = "Top 10 GO Biological Process Enrichment results"))
  
  # Try to add plot
  go_plot <- find_plot(res_dir, ".*_go_barplot\\.png$")
  if (!is.null(go_plot)) {
      cat("\n\n### Functional Visualization (GO)\n")
      cat(glue("![GO Enrichment Visualization]({go_plot}){{width=80% fig-pos='H'}}\n\n"))
  }
} else {
  cat("No significant GO enrichment observed.\n")
}
```

## KEGG Pathway Summary
```{r kegg-results, results='asis'}
if(nrow(kegg) > 0 && !"Message" %in% colnames(kegg)) {
  cols <- intersect(colnames(kegg), c("ID", "Description", "pvalue", "p.adjust"))
  cols_present <- intersect(cols, colnames(kegg))
  top_kegg <- kegg[order(kegg$pvalue), ]
  top_kegg <- head(top_kegg[, cols_present, drop = FALSE], 10)
  print(kable_report(top_kegg, caption = "Top 10 KEGG Pathway Enrichment results"))
  
  kegg_plot <- find_plot(res_dir, ".*_kegg_barplot\\.png$")
  if (!is.null(kegg_plot)) {
      cat("\n\n### Pathway Visualization (KEGG)\n")
      cat(glue("![KEGG Pathway Visualization]({kegg_plot}){{width=80% fig-pos='H'}}\n\n"))
  }
  
  # Try to find and add Pathview molecular pathway map
  pathview_plot <- find_plot(res_dir, ".*pathview\\.png$")
  if (!is.null(pathview_plot)) {
      cat("\n\n### Molecular Pathway Diagram (Pathview)\n")
      cat(glue("![KEGG Molecular Pathway Map]({pathview_plot}){{width=95% fig-pos='H'}}\n\n"))
  }
} else {
  cat("No significant KEGG pathway enrichment observed.\n")
}
```


# Disease Ontology Analysis

```{r disease-results, results='asis'}
if(nrow(disease) > 0 && !"Message" %in% colnames(disease)) {
  cols <- intersect(colnames(disease), c("ID", "Description", "pvalue", "p.adjust"))
  cols_present <- intersect(cols, colnames(disease))
  top_disease <- disease[order(disease$pvalue), ]
  top_disease <- head(top_disease[, cols_present, drop = FALSE], 10)
  print(kable_report(top_disease, caption = "Significant Disease-Gene Associations"))
  
  dis_plot <- find_plot(res_dir, ".*_disease_barplot\\.png$")
  if (!is.null(dis_plot)) {
      cat("\n\n### Disease Association Visualization\n")
      cat(glue("![Disease Enrichment Visualization]({dis_plot}){{width=80% fig-pos='H'}}\n\n"))
  }
} else {
  cat("No significant Disease Ontology enrichment observed.\n")
}
```

\newpage

# Pipeline Traceability & Methods

## Methods

```{r results='asis'}
# Capture the methods description from run_info
meth_description_data <- if (!is.null(run_info$methods_description)) run_info$methods_description else ""

if (meth_description_data != "") {
    # Perform a robust transformation of HTML description to Markdown
    # to avoid raw HTML rendering issues in both PDF and HTML formats
    m_text <- meth_description_data
    m_text <- gsub("<h4>", "\n### ", m_text)
    m_text <- gsub("</h4>", "\n", m_text)
    m_text <- gsub("<p>", "\n\n", m_text)
    m_text <- gsub("</p>", "\n", m_text)
    m_text <- gsub("<li>", "\n* ", m_text)
    m_text <- gsub("</li>", "", m_text)
    m_text <- gsub("<strong>|<b>", "**", m_text)
    m_text <- gsub("</strong>|</b>", "**", m_text)
    m_text <- gsub("<em>|<i>", "*", m_text)
    m_text <- gsub("</em>|</i>", "*", m_text)
    
    # Strip any remaining HTML tags
    m_text <- gsub("<[^>]*>", "", m_text)
    
    # Clean up whitespace: trim and ensure no leading spaces per line that trigger code blocks
    m_text <- unlist(strsplit(m_text, "\n"))
    m_text <- trimws(m_text)
    meth_text <- paste(m_text, collapse = "\n")
    
    # Collapse multiple newlines
    meth_text <- gsub("\n{3,}", "\n\n", meth_text)
    
    # Output as-is with a trailing newline for Pandoc safety
    cat(meth_text, "\n")
} else {
    cat("Detailed descriptions of analytical methods were not captured for this run.\n")
}
```

# References

::: {#refs}
:::

\newpage

# Appendix: Data Audit & Traceability

## Analytical Software versions

The following software versions were utilised for this analysis. For a full technical audit log, including configuration parameters and alignment statistics, please consult the primary **MultiQC Report**.

```{r results='asis'}
if (length(software_versions) > 0) {
    sv_df <- data.frame(
        Tool = names(software_versions),
        Version = unlist(software_versions)
    )
    kable_report(sv_df, caption = "Software Traceability")
} else {
    cat("*Software version data was not captured for this specific run. All tools are maintained within the pipeline's reproducible containerized environment. Primary diagnostics are recorded in the MultiQC report.*\n\n")
}
```

## Pipeline Run Information
```{r results='asis'}
run_name <- if (!is.null(run_info$run_name)) run_info$run_name else "Unknown"
run_version <- if (!is.null(run_info$version)) run_info$version else "1.2.0"
run_time <- if (!is.null(run_info$timestamp)) run_info$timestamp else Sys.time()
run_files <- if (!is.null(run_info$files_processed)) run_info$files_processed else 0
run_commit <- if (!is.null(run_info$commit_hash)) run_info$commit_hash else "Not available"
run_cmd <- if (!is.null(run_info$command_line)) run_info$command_line else "Not available"
run_nf <- if (!is.null(run_info$nextflow_version)) run_info$nextflow_version else "Not available"
run_ce <- if (!is.null(run_info$container_engine)) run_info$container_engine else "Local/Native"
run_genome <- if (!is.null(run_info$genome_build)) run_info$genome_build else "Custom/Unspecified"

cat(glue("- **Run Name:** {run_name}\n\n"))
cat(glue("- **Execution UUID:** UUID_PLACEHOLDER\n\n"))
cat(glue("- **Pipeline Version:** {run_version}\n\n"))
cat(glue("- **Nextflow Engine:** {run_nf}\n\n"))
cat(glue("- **Container Engine:** {run_ce}\n\n"))
cat(glue("- **Reference Genome:** {run_genome}\n\n"))
cat(glue("- **Commit Hash:** {run_commit}\n\n"))
cat(glue("- **Command Line:** {run_cmd}\n\n"))
cat(glue("- **Timestamp:** {run_time}\n\n"))
cat(glue("- **Total Files Processed:** {run_files}\n\n"))

# Add Enrichment Parameters
if (!is.null(run_info$enrichment_params)) {
    e_p <- run_info$enrichment_params
    pv <- if (!is.null(e_p$pvalue_cutoff)) e_p$pvalue_cutoff else "0.05"
    lfc <- if (!is.null(e_p$logfc_cutoff)) e_p$logfc_cutoff else "0.5"
    top <- if (!is.null(e_p$top_n_genes)) e_p$top_n_genes else "100"

    cat("\n### Enrichment Cutoffs\n\n")
    cat(glue("- **P-value Threshold:** {pv}\n\n"))
    cat(glue("- **LogFC Threshold:** {lfc}\n\n"))
    cat(glue("- **Top Genes Prioritized:** {top}\n\n"))
}

if (!is.null(run_info$sanity_checks)) {
    cat("\n### Output Sanity Checks\n\n")
    for (check in run_info$sanity_checks) {
        cat(glue("- {check}\n"))
    }
    cat("\n")
}

if (!is.null(run_info$checksums) && length(run_info$checksums) > 0) {
    cat("\n### Input Integrity (SHA256)\n\n")
    for (fname in names(run_info$checksums)) {
        cat(glue("- **{fname}**: `{run_info$checksums[[fname]]}`\n"))
    }
    cat("\n")
}

if (!is.null(run_info$resolved_params) && length(run_info$resolved_params) > 0) {
    cat("\n### Resolved Nextflow Parameters\n\n")
    cat("```json\n")
    cat(toJSON(run_info$resolved_params, pretty=TRUE, auto_unbox=TRUE))
    cat("\n```\n\n")
}
```

## Datapath Audit
```{r audit, results='asis'}
cat("Verification of input result paths used for this report generation.\n\n")
audit_data <- function(name, df, path) {
    status <- if(nrow(df) > 0) "SUCCESS" else "NO DATA"
    if (knitr::is_html_output()) {
        cat(glue("- **{name}**: {status} ({nrow(df)} records) \n    *(Source: {path})*\n"))
    } else {
        cat(glue("- **{name}**: {status} ({nrow(df)} records) \n    *(Source: \\path{{{path}}})*\n"))
    }
}

audit_data("Samples", meta, params$sample_metadata)
audit_data("QC Summary", qc, params$qc_summary)
audit_data("DMRs", dmr, params$dmr_summary)
audit_data("Prioritized Genes", gene, params$gene_summary)
audit_data("GO Enrichment", pathways, params$pathway_summary)
audit_data("KEGG Enrichment", kegg, params$kegg_summary)
audit_data("Disease Enrichment", disease, params$disease_summary)
```

\vfill
\vspace{2cm}
\hrule
\vspace{2mm}
\begin{center}
\small \textit{Report generated automatically for research purposes only. Analytical values should be validated with primary experimental findings and the MultiQC quality report.}
\end{center}

